---
title: "Reading List: AI Alignment & LLM Benchmarking"
author: "Research Companion"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(DT)
library(dplyr)
library(jsonlite)
library(RefManageR)
```

<style>
body {
  font-family: system-ui, -apple-system, 'Segoe UI', Roboto, 'Helvetica Neue', Arial, sans-serif;
  color: #1a1a1a;
  background-color: #F7FAFC;
}
.main-container {
  max-width: 1100px !important;
  margin: 40px auto;
  background: white;
  padding: 40px;
  border-radius: 12px;
  box-shadow: 0 4px 6px rgba(0, 0, 0, 0.05);
}
.readme-content {
  font-family: system-ui, -apple-system, 'Segoe UI', Roboto, 'Helvetica Neue', Arial, sans-serif;
  line-height: 1.6;
}
.info-box {
  border-left: 4px solid darkblue;
  background: #f0f4ff;
  padding: 1em 1.25em;
  border-radius: 4px;
  margin: 1em 0;
}
.info-box strong {
  color: darkblue;
}
.section-intro {
  font-style: italic;
  color: #333;
  padding: 0.5em 1em;
  border-left: 2px solid #8ab4f8;
  background: #f8faff;
  border-radius: 3px;
  margin: 0.5em 0 1.5em 0;
}
table.dataTable tbody tr {
  cursor: pointer;
  transition: background-color 0.2s ease;
}
table.dataTable tbody tr:hover {
  background-color: #EDF2F7 !important;
}
td.details-control {
  cursor: pointer;
}
.citation-details {
  background: #F8FAFC;
  padding: 20px;
  border-left: 4px solid darkblue;
  border-radius: 0 8px 8px 0;
  margin: 10px 0;
}
.details-title {
  font-weight: 600;
  color: darkblue;
  margin-top: 0;
  margin-bottom: 8px;
}
.details-abstract {
  font-style: italic;
  color: #4A5568;
  line-height: 1.6;
}
.details-meta {
  margin-top: 12px;
  font-size: 0.9em;
  color: #718096;
}
</style>

<div class="readme-content">

# Reading List: <span style="color: darkblue; font-weight: 500;">AI Alignment</span> & LLM Benchmarking

<div class="section-intro">
A curated collection of publications on <span style="color: darkblue; font-weight: 500;">large language model</span> benchmarking and evaluation, selected to support the <span style="color: darkblue; font-weight: 500;">AI Alignment</span> workshop series. Topics span evaluation frameworks, benchmark design and critique, multilingual and domain-specific assessments, educational applications, responsible AI, and the environmental impact of LLM inference — reflecting the breadth of the field and its relevance to <span style="color: darkblue; font-weight: 500;">Dutch higher education</span>.
</div>

<div class="info-box">
<strong>How to use this reading list</strong><br>
Click the <strong>DOI</strong> link in the first column to open the original paper. Use the search box to filter by author, title, or keyword.
</div>

```{r process-bib}
# 1. Read and parse the external BibTeX database
bib_data <- ReadBib("references.bib", check = FALSE)
bib_df <- as.data.frame(bib_data)

# 2. Safely ensure all possible columns exist to avoid compilation crashes
required_cols <- c("author", "year", "title", "journal", "booktitle", "abstract", "keywords", "url", "doi")
for (col in required_cols) {
  if (!col %in% names(bib_df)) {
    bib_df[[col]] <- NA_character_
  }
}

# 3. Normalize columns securely now that columns are guaranteed to exist
bib_clean <- bib_df %>%
  mutate(
    Author = ifelse(is.na(author), "Unknown Author", as.character(author)),
    Year = ifelse(is.na(year), "N/A", as.character(year)),
    Title = ifelse(is.na(title), "Untitled", as.character(title)),
    Journal = case_when(
      !is.na(journal) ~ as.character(journal),
      !is.na(booktitle) ~ as.character(booktitle),
      TRUE ~ "Other Source"
    ),
    Abstract = ifelse(is.na(abstract), "No abstract provided for this reference.", as.character(abstract)),
    Keywords = ifelse(is.na(keywords), "None", as.character(keywords)),
    Link = ifelse(is.na(url) | url == "", "", paste0('<a href="', url, '" target="_blank">', doi, '</a>')),
    ChildHTML = paste0(
      '<div class="citation-details">',
        '<h4 class="details-title">', Title, '</h4>',
        '<p class="details-abstract"><strong>Abstract:</strong> ', Abstract, '</p>',
        '<div class="details-meta"><strong>Keywords:</strong> ', Keywords, '</div>',
      '</div>'
    )
  )

# 4. Create the visible skeleton table columns
display_df <- bib_clean %>%
  select(Link, Author, Year, Title, Journal)
```

```{r render-table}
# 4. Bind data frame to an interactive datatable view
datatable(
  display_df,
  escape = FALSE,
  options = list(
    columnDefs = list(
      list(className = 'details-control', targets = 3)
    ),
    pageLength = 10,
    dom = 'ftip' # Enables searching (f), table (t), info (i), and pagination (p)
  ),
  callback = JS(paste0(
    "var raw_data = ", jsonlite::toJSON(bib_clean), ";",
    "table.on('click', 'td.details-control', function() {",
    "  var tr = $(this).closest('tr');",
    "  var row = table.row(tr);",
    "  if (row.child.isShown()) {",
    "    row.child.hide();",
    "    tr.removeClass('shown');",
    "  } else {",
    "    var idx = row.index();",
    "    var record = raw_data[idx];",
    "    row.child(record.ChildHTML).show();",
    "    tr.addClass('shown');",
    "  }",
    "});"
  ))
)
```

---

<p style="color: #555;"><em>Last updated: `r Sys.Date()` — references curated for the AI Alignment workshop materials.</em></p>

</div>
